# Adding datasets (brief)

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = "asis",
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

library(traits.build)

my_kable_styling <- util_kable_styling_html
```

```{r, echo=FALSE, results='hide', message=FALSE}
## Loads austraits into global name space
austraits <- austraits:::austraits_lite

schema <- get_schema()
definitions <- austraits$definitions
```

This section gives a brief overview of how to add datasets to the database. For a more detailed guide, see [adding_datasets](adding_data_long.html).

It is important that all steps are followed so that our automated workflow proceeds without problems.

1. Clone the `traits.build` repository from github
2. Create a new branch in the repo, named for the new `dataset_id` in `author_year` format, e.g. `Gallagher_2014`.
3. Create a new folder within the folder `data` with the name `dataset_id`, e.g. `Gallagher_2014`.
4. Prepare the file `data.csv` and place it within the new folder ([details](#csv_file) here).
5. Prepare the file `metadata.yml` and place it within the new folder ([details](#metadata_file) here).
6. Add the new study into the build framework and rebuild AusTraits, by running `build_setup_pipeline()`.

This step updates the file `remake.yml` with appropriate rules for the new dataset; similarly if you remove datasets, do the same. (At this stage, [remake](https://github.com/richfitz/remake) offers no looping constructs, so for now we generate the remake file using: [whisker](https://github.com/edwindj/whisker).)

You can then rebuild AusTraits, including your dataset.

7. Run tests and quality checks on the newly added dataset and correct the `data.csv` and `metadata.yml` files as necessary.
8. Generate and proofread a report on the data. In particular, check that numeric trait values fall within a logical range relative to other studies, and that individual trait observations are not unnecessarily excluded because their trait values are unsupported.
9. Return to step 6 if changes are made to the `data.csv` or `metadata.yml` files.
10. Push the GitHub branch.

The best place to get started learning how to add datasets is to work through a series of 7 [tutorials](tutorial_dataset_1.html). Each introduces you to specific [traits.build functions](https://traitecoevo.github.io/traits.build/reference/index.html) designed to facilitate the addition of dataset metadata or the metadata formats for specific types of datasets. 

It may also help to download one of the [existing datasets](https://github.com/traitecoevo/austraits.build/tree/master/data) to use as a template for your own files and a guide on required content. 

You should look at the files in the [config folder](https://github.com/traitecoevo/austraits.build/tree/master/config), particularly the `definitions` file for the list of traits in AusTraits and how trait definitions are formatted.

The exhaustive [adding data](adding_data_long.html) section offers a fairly comprehensive run through of each step and all the known variants on how to fill the various metadata fields, but is likely overwhelming until you are familiar with the traits.build workflow and metadata format.
