# AusTraits, Australia's Plant Trait Database

[AusTraits](https://github.com/traitecoevo/austraits.build), Australia's largest plant trait database was the first trait database to be build using the traits.build workflow.

## AusTraits data records

As of October, 2023, AusTraits has:

-   370+ datasets 

-   250+ contributors 

-   1,800,000+ data records 

-   500+ traits 

-   30,000 Australian plant taxa 

## Accessing AusTraits data

There are multiple ways to access and manipulate AusTraits data.\

-   Download the dataset from [Zenodo](https://doi.org/10.5281/zenodo.3568417)\

-   Source AusTraits using the [austraits R package](https://github.com/traitecoevo/austraits)\

-   Clone the austraits.build GitHub repository and build AusTraits from the current [develop branch](https://github.com/traitecoevo/austraits.build) (*not easily reproducible*)

### Download database with austraits.R package

First install and source the package: 

```{r, eval = FALSE}
install.packages("remotes")  # if it isn't already installed on your machine

remotes::install_github("traitecoevo/austraits", dependencies = TRUE, upgrade = "ask")

library(austraits) 
```

Before loading AusTraits, see what versions are available,

```{r  eval = FALSE}
get_versions()
```

yields:

```{r, eval = FALSE}
# A tibble: 6 × 3
  date       version doi                   
  <chr>      <chr>   <chr>                 
1 2023-01-30 4.2.0   10.5281/zenodo.8353840 
2 2023-01-30 4.1.0   10.5281/zenodo.7583087
3 2022-11-27 4.0.0   10.5281/zenodo.7368074
4 2021-07-14 3.0.2   10.5281/zenodo.5112001
5 2021-07-14 3.0.1   10.5281/zenodo.5099552
6 2021-07-14 3.0.0   10.5281/zenodo.5099545
```

Download the most recently released version:

```{r, eval = FALSE}
austraits <- austraits::load_austraits(doi = "10.5281/zenodo.8353840") # you can load from the Zenodo doi

austraits <- austraits::load_austraits(version = "4.1.0") # you can load from the version
                                                          # the `path` argument stores a copy for future use.
```