# AusTraits, Australia's Plant Trait Database

[AusTraits](https://github.com/traitecoevo/austraits.build), Australia's largest plant trait database was the first trait database to be build using the traits.build workflow.

## AusTraits data records

As of October, 2023, AusTraits has:

-   370+ datasets 

-   250+ contributors 

-   1,800,000+ data records 

-   500+ traits 

-   30,000 Australian plant taxa 

## Accessing AusTraits data

There are multiple ways to access and manipulate AusTraits data.\

-   Download the dataset from [Zenodo](https://doi.org/10.5281/zenodo.3568417)\

-   Source AusTraits using the [austraits R package](https://github.com/traitecoevo/austraits)\

-   Clone the austraits.build GitHub repository and build AusTraits from the current [develop branch](https://github.com/traitecoevo/austraits.build) (*not easily reproducible*)

### Download database with austraits.R package

First install and source the package: 

```{r, eval = FALSE}
install.packages("remotes")  # if it isn't already installed on your machine

remotes::install_github("traitecoevo/austraits", dependencies = TRUE, upgrade = "ask")

library(austraits) 
```

Before loading AusTraits, see what versions are available,

```{r  eval = TRUE}
austraits::get_versions()
```

Download the most recently released version:

```{r, eval = FALSE}
most_recent_doi <- austraits::get_versions()|> dplyr::pull("doi") |> dplyr::first()

most_recent_doi

austraits <- austraits::load_austraits(doi = most_recent) # you can load from the Zenodo doi

austraits <- austraits::load_austraits(version = "5.0.0") # you can load from the version
                                                          # the `path` argument stores a copy for future use.
```