# Troubleshooting common issues

This chapter collects the errors and warnings you are most likely to meet while adding a dataset and
compiling a `traits.build` database, and explains how to resolve each one. It is the "detailed
troubleshooting" companion to the [Getting help](help.html) chapter: work through the relevant
section here before posting a question.

Most problems fall into a few families:

- [Categorical trait values that aren't recognised](#unsupported-values)
- [Numeric values outside the allowable range](#out-of-range)
- [Taxon names that don't align](#taxonomy)
- [A dataset that won't pivot to wide format](#cannot_pivot)
- [Duplicate measurements](#duplicates)
- [Errors reading `metadata.yml` or `traits.yml`](#yaml-errors)

Two general habits solve a large share of issues. First, read the error message carefully — the
`traits.build` messages are designed to name the offending trait, dataset, or value. Second, before
assuming the package is at fault, check whether the problem lies in your data files or custom R code;
see [attribute the error](help.html#sec-contact) in the Getting help chapter.

## Unsupported trait values {#unsupported-values}

This error occurs when, for a categorical trait, a value in [data.csv]{style="color:blue;"} is not
one of the allowable values defined for that trait in the trait dictionary
([config/traits.yml]{style="color:blue;"}). The offending records are moved to the `excluded_data`
table rather than silently dropped.

List the values that need a substitution:

```{r, eval=FALSE}
table <- my_database$excluded_data %>%
  filter(dataset_id == current_study) %>%
  filter(error == "Unsupported trait value") %>%
  select(dataset_id, trait_name, value) %>%
  distinct()
```

You then map each unrecognised value onto an allowable one using a *substitution*. Add a single
substitution to [metadata.yml]{style="color:blue;"} with `metadata_add_substitution()`:

```{r, eval=FALSE}
metadata_add_substitution(
  dataset_id = current_study,
  trait_name = "plant_growth_form",
  find = "T",
  replace = "tree"
)
```

When many values need fixing, edit the table above (add a `replace` column) and read the whole set in
at once with `metadata_add_substitutions_table()`. The table must have the columns
[dataset_id]{style="color:blue;"}, [trait_name]{style="color:blue;"}, [find]{style="color:blue;"}, and
[replace]{style="color:blue;"}:

```{r, eval=FALSE}
table <- table %>%
  rename(find = value) %>%
  mutate(replace = c("tree", "mallee", "shrub", "graminoid", "herb"))

metadata_add_substitutions_table(
  table,
  dataset_id = dataset_id,
  trait_name = trait_name,
  find = find,
  replace = replace
)
```

For a long list it is often easiest to write the table to a CSV, edit it in a spreadsheet or text
editor, and read it back:

```{r, eval=FALSE}
write_csv(table, "data/dataset_id/raw/substitutions_required.csv")

# ...edit outside of R...

table <- read_csv("data/dataset_id/raw/substitutions_required.csv")
```

See [adding substitutions](adding_data_long.html#add_substitutions) for the full workflow, and
`dataset_check_categorical_substitutions()` in the
[data-quality functions chapter](check_dataset_functions.html) for a helper that builds the table for
you.

## Numeric values out of range {#out-of-range}

For a numeric trait, a value that falls outside the allowable `range` defined for that trait in
[config/traits.yml]{style="color:blue;"} is moved to `excluded_data`. Inspect the excluded values:

```{r, eval=FALSE}
my_database$excluded_data %>%
  filter(dataset_id == current_study,
         error == "Value out of allowable range") %>%
  select(dataset_id, trait_name, value, observation_id, unit, original_name)
```

If the list is long, the cause is almost always a **units-conversion problem** — the incoming units
in the metadata do not match what the data actually contains, so every value is scaled wrongly. Check
the `unit_in` you mapped for the trait against the units in the source, and confirm the conversion
exists in [config/unit_conversions.csv]{style="color:blue;"}. If only a few values are excluded, they
may be legitimately out of range; confirm they are genuine errors before dismissing them.
`dataset_check_numeric_values()` in the
[data-quality functions chapter](check_dataset_functions.html) tabulates these for review.

## Taxon names that don't align {#taxonomy}

An `aligned_name` in a newly added dataset may not appear in the database's
[config/taxon_list.csv]{style="color:blue;"} for two reasons: the name needs aligning (a typo or
non-standard syntax), or the taxon list simply doesn't yet include that name and must be extended
from an external taxonomic resource.

Each database maintains its own taxonomy — the taxon list must be built outside the `traits.build`
workflow because different databases reference different taxonomic sources (AusTraits, for example,
uses the Australian Plant Census and the Australian Plant Names Index). Generate the list of names
still needing attention with `dataset_check_taxonomic_updates()` (see the
[data-quality functions chapter](check_dataset_functions.html)), then either add taxonomic updates to
the dataset's [metadata.yml]{style="color:blue;"} (see
[adding taxonomic updates](adding_data_long.html#add_taxonomic_updates)) or extend the taxon list.

## Dataset can't pivot to wide format {#cannot_pivot}

To convert a `traits.build` database to wide format, the `traits` table must be able to pivot wider.
One of the tests run by `dataset_test()` — `traits.build::check_pivot_wider()` — confirms this. It
checks that every row in the `traits` table has a **unique combination of seven columns**:

`dataset_id`, `trait_name`, `observation_id`, `value_type`, `repeat_measurements_id`, `method_id`,
and `method_context_id`.

If two rows share the same combination, the table cannot pivot and the test fails. In almost every
case the real cause is that `observation_id` has not been generated as you intended — two genuinely
distinct observations were assigned the same `observation_id` because a distinguishing piece of
metadata was not mapped. (See the [glossary](glossary.html#observations) for how `observation_id`
is built.)

There are two common explanations and solutions:

1.  **Species-level values repeated alongside individual-level measurements.** If your dataset mixes
    individual- (or population-) level measurements with species-level measurements, the same
    species-level value may be read in on many rows. Retain only the first instance of each
    species-level value using `custom_R_code`, where `taxon_name` is the column of taxon names and
    `column 1`, `column 2`, … are the categorical-trait columns that need de-duplicating:

```{r, eval=FALSE}
data %>%
  group_by(taxon_name) %>%
  mutate(across(c("column 1", "column 2", "column 3"), replace_duplicates_with_NA)) %>%
  ungroup()
```

2.  **Repeat measurements over time (or by method) not distinguished.** Rows representing
    measurements made at different times, or with different methods, on the same entity need a
    `temporal_context_id` / `repeat_measurements_id` (for repeat measurements over time) or a
    `method_context_id` (for multiple methods) so that they resolve to distinct rows. Map the column
    that distinguishes them, or add the appropriate context, so each measurement is uniquely
    identified.

To find exactly which measurements are blocking the pivot, run `dataset_check_not_pivoting()` (see
the [data-quality functions chapter](check_dataset_functions.html)); it returns the specific rows with
duplicate combinations so you can trace the missing piece of metadata.

## Duplicate measurements {#duplicates}

Duplicates come in two forms.

**Within a dataset**, the same taxon × trait value can legitimately appear many times — for example a
species-level value (like `plant_growth_form`) reported on every individual's row, or a single bulked
measurement reported against every contributing individual. When such repetition is unintended, remove
it with `custom_R_code` (see [custom R code](adding_data_long.html#custom_R)). The helper
`dataset_check_duplicates_within_dataset()` flags suspicious within-dataset duplication — be alert
when `n_duplicates` is large or is identical across every taxon for a trait.

**Across datasets**, the same measurements may be submitted twice — for instance when two
collaborators contribute the same data, or when a dataset arrives both from its collector and inside a
broader compilation. The automated tests eliminate *most* exact duplicates, but cross-dataset
detection is genuinely hard (it depends on how many significant figures to compare, and traits with
naturally narrow ranges produce false positives). See
`dataset_check_duplicates_across_datasets()` in the
[data-quality functions chapter](check_dataset_functions.html) for the current status of this check.

## Errors reading `metadata.yml` or `traits.yml` {#yaml-errors}

Both the dataset metadata files and the trait dictionary are [YAML](yaml.html) files, where meaning is
carried entirely by **indentation** and structure. Small formatting slips produce errors that can look
alarming but are quick to fix. The most common are:

- **`incomplete final line found on '...'`** — the file is missing a trailing newline, or has stray
  spaces or blank lines at the end. Open the file in a proper text editor, remove trailing whitespace,
  and ensure the file ends with a single newline.
- **Wrong indentation.** Copying a block from one trait or dataset into another often carries the
  wrong indentation, which changes the structure YAML infers. Indent with **spaces, never tabs**, and
  keep sibling entries aligned to the same column.
- **Unquoted special characters.** Values containing `:`, `#`, `%`, leading/trailing brackets, or a
  leading `>` or `|` can be misread. Wrap such values in quotes — for example units like
  `"umol/m2/s"` or a description containing a colon.

The [creating a trait dictionary](create_dictionary.html) chapter has a dedicated section on writing
valid `traits.yml` files and the formatting pitfalls to watch for, including a worked example of an
indentation error and how to fix it. A useful habit is to validate a file by reading it back with
`yaml::read_yaml("config/traits.yml")` — if it parses without error, the structure is sound.

## Still stuck?

If none of the above resolves your problem, first update to the latest `traits.build` release in case
the issue is a bug that has since been fixed, then follow the [Getting help](help.html) chapter:
attribute the error, prepare a minimal reproducible example, and post to the relevant GitHub
Discussions page.
