# Trait dictionary

A core component of the traits.build database structure is a trait dictionary that clearly defines and describes each trait concept included within the trait database.

A description of the trait dictionary is available at: Wenk, E.H., Sauquet, H., Gallagher, R.V., Brownlee, R., Boettiger, C., Coleman, D., Yang, S., Auld, T., Barrett, R., Brodribb, T., Choat, B., Dun, L., Ellsworth, D., Gosper, C., Guja, L., Jordan, G.J., Le Breton, T., Leigh, A., Lu-Irving, P., Medlyn, B., Nolan, R., Ooi, M., Sommerville, K.D., Vesk, P., White, M., Wright, I.J., Falster, D.S., 2024. The AusTraits plant dictionary. Sci Data 11, 537. [DOI: 10.1038/s41597-024-03368-z](https://doi.org/10.1038/s41597-024-03368-z).


Building upon the standard outlined by the ETS [Ecological Trait-data Standard](https://doi.org/10.1111/2041-210X.13288), all numeric traits must have standard units and an allowable range. Categorical traits must have a list of allowable trait values.

Each trait concept must have a complete, explicit definition to eliminate confusion between trait concepts.

The trait dictionary is compiled in a simple yml format, allowing trait definitions to be easily added and edited.

## Examples

A sample numeric trait definition:

``` {r, eval = FALSE}
seed_dry_mass:
    label: Seed dry mass
    description: A seed morphology trait [TO:0000184] which is the dry [PATO:0001824]
    mass [PATO:0000125] of a mature [PATO:0001701] seed [PO:0009010].;Dry mass
    of a mature seed, including both oven dried and air-dried samples.
    comments: Standard methods people will have used to dry seeds include, 'fresh'
    (at dispersal, mature); 'air dried' (at local ambient conditions); 'seed bank
    air dried' (to 15% relative humidity); and 'oven dried' (>100 deg C for a
    set number of hours; e.g. seed bank standard is 103 deg C for 17 hours). It
    is expected that some observations in AusTraits mapped onto ‘seed_dry_mass'
    will actually include both the seed and some dispersal tissue, if the two
    cannot easily be separated; these should be mapped to 'diaspore_dry_mass'.
    type: numeric
    units: mg
    allowed_values_min: 1.0e-05
    allowed_values_max: 1000000.0
```

A sample categorical trait definition:

``` {r, eval = FALSE}
leaf_compoundness:
    label: Leaf compoundness
    description: A leaf shape trait [TO:0000492] which is whether a leaf [PO:0025034]
    is a simple leaf [PO:0020042] or is divided into leaflets [PO:0020049] making
    it a compound leaf [PO:0020043].;A binary trait that indicates whether a leaf
    lamina is simple or divided into discontinuous leaflets (compound).
    comments: See also the trait 'leaf_division' for more detailed leaf compoundness
    trait values. Note that there might be some species that grade from deeply
    lobed simple leaves (see trait `leaf_lobation`)  to compound leaves.
    type: categorical
    allowed_values_levels:
        compound: A leaf that is divided into multiple leaflets. [PO:0020043]
        simple: A leaf with a single undivided blade. [PO:0020042]
```

For additional examples, see the complete dictionary used for the [AusTraits database](https://github.com/traitecoevo/austraits.build/tree/develop/config).

