# Installation and quick start

This chapter gets you from a clean R installation to a working `traits.build` database in a few
minutes. It is the fastest way to see what a compilation looks like before reading the concept
chapters that follow. If you would like a fuller, step-by-step walk-through, see the
[example compilation tutorial](tutorial_compilation.html) and the
[guide to adding data](tutorial_datasets.html).

## What you need

- **R** (and, we recommend, RStudio). `traits.build` has no dependencies on proprietary software.
- The **`remotes`** package, for installing from GitHub: `install.packages("remotes")`.
- A **Git client** to clone the template. [GitHub Desktop](https://desktop.github.com/) is a
  friendly option if you are new to Git; RStudio's built-in Git support or the command line work
  equally well. If Git and GitHub are new to you, the free book
  [Happy Git with R](https://happygitwithr.com/) is an excellent starting point.

## Step 1 — Install the `traits.build` package

`traits.build` is the engine that compiles a database. Install the current release from GitHub:

```{r, eval=FALSE}
install.packages("remotes")  # if not already installed
remotes::install_github("traitecoevo/traits.build", dependencies = TRUE)
```

For the most up-to-date installation notes — including how to install a specific version — see the
[package installation instructions](https://traitecoevo.github.io/traits.build/index.html#installation).

## Step 2 — Clone the template

[`traits.build-template`](https://github.com/traitecoevo/traits.build-template) is a ready-to-build
example repository. It contains everything a compilation needs — a sample
[trait dictionary](https://github.com/traitecoevo/traits.build-template/blob/master/config/traits.yml),
a [unit-conversions file](https://github.com/traitecoevo/traits.build-template/blob/master/config/unit_conversions.csv),
a [database metadata file](https://github.com/traitecoevo/traits.build-template/blob/master/config/metadata.yml),
a placeholder [taxon list](https://github.com/traitecoevo/traits.build-template/blob/master/config/taxon_list.csv),
and a script of reusable
[custom R functions](https://github.com/traitecoevo/traits.build-template/blob/master/R/custom_R_code.R) —
plus two datasets already added so you can build a working database immediately, and further datasets
with tutorials that introduce the workflow.

Clone it by visiting the [repository](https://github.com/traitecoevo/traits.build-template), clicking
the green **`<> Code`** button and choosing your preferred method, or from the command line:

```bash
git clone https://github.com/traitecoevo/traits.build-template.git
```

## Step 3 — Build the example database

Open the cloned project in R (set the working directory to the repository root), then generate the
build script and run it:

```{r, eval=FALSE}
library(traits.build)

build_setup_pipeline(method = "base", database_name = "traits.build_database")
source("build.R")
```

`build_setup_pipeline()` writes a `build.R` script tailored to your compilation; sourcing it reads
the datasets in `data/`, harmonises them against the trait dictionary, and assembles the relational
database. When it finishes you have a `traits.build` database object in your R session.

## Step 4 — Look around

The compiled database is a list of linked tables. Explore its structure:

```{r, eval=FALSE}
names(traits.build_database)

# the core long-format table of trait measurements
traits.build_database$traits

# metadata and provenance for the build
traits.build_database$build_info
```

The [data structure chapter](database_structure.html) explains each of these components, and the
[glossary](glossary.html) defines the identifiers that link them.

## Where to go next

- To understand *why* the database is shaped this way, read [Motivation](motivation.html) and
  [Workflow](workflow.html).
- To add your own data, work through the [guide to adding data](tutorial_datasets.html), which walks
  through seven progressively more complex datasets.
- If you hit an error, the [troubleshooting chapter](data_common_issues.html) covers the most common
  problems, and [Getting help](help.html) explains how to ask a good question.

::: {.callout-note}
The template's build pipeline currently uses [`remake`](https://github.com/richfitz/remake), which is
no longer actively maintained; the project expects to move to
[`targets`](https://docs.ropensci.org/targets/) in a future release. The `build_setup_pipeline()` /
`build.R` workflow shown above shields you from that change — follow the template's own README for
the current build command if it has been updated.
:::
