
# Analysis example: Using AusTraits with spatial data

This tutorial is based on [this great tutorial](https://labs.ala.org.au/posts/2023-08-28_alternatives-to-box-plots/post.html) by Dax Kellie and Shandiya Balasubramaniam from the ALA team.


```{r, eval=TRUE, message = FALSE, warning = FALSE}
# remotes::install_github("traitecoevo/austraits")
library(tidyverse)
library(austraits)
```


Load austraits

```{r, eval = TRUE}
most_recent <- austraits::get_versions()[["doi"]][1]

most_recent

austraits <- austraits::load_austraits(doi = most_recent)
```
Extract leaf mass per area (LMA) data

```{r, eval = TRUE}
# You can use `lookup_trait()` to search for traits containing keywords
austraits::lookup_trait(austraits, "leaf_mass")

# Get trait data
leaf_mass <- austraits %>%
  austraits::extract_trait("leaf_mass_per_area") %>%
  purrr::pluck("traits") # Grab the traits table from the list of austraits tables
```


Filter to six species in the dataset

```{r, eval = TRUE}
sample_names <- c("Cryptocarya rigida", "Pteridium esculentum",
                  "Eucalyptus baxteri", "Melaleuca armillaris",
                  "Eucalyptus wandoo", "Eucalyptus piperita")

leaf_mass_sample <- leaf_mass %>% dplyr::filter(taxon_name %in% sample_names)
```


Plot raincloud plot of LMA for the six species

```{r, eval = F}
# install.packages(c("ggdist", "gghalves", "ggtext"))
# remotes::install_github("olihawkins/pilot")
library(ggplot2)
library(ggdist)
library(gghalves)
library(ggtext)
library(pilot)

ggplot2::ggplot(
  data = leaf_mass_sample,
  aes(x = taxon_name %>% stringr::str_wrap(10) %>% reorder(value),
      y = value,
      colour = taxon_name,
      fill = taxon_name)
) +
  ggdist::stat_halfeye(
    adjust = .4,
    width = .87,
    colour = NA) +
  gghalves::geom_half_point(
    side = "l",
    range_scale = .3,
    alpha = .6,
    size = 2.2) +
  geom_boxplot(
    aes(colour = taxon_name,
        colour = after_scale(colorspace::darken(colour, .7))),
    width = .12, # Adjust box width
    fill = NA,
    size = 1.1, # Size of box line
    outlier.shape = NA # Remove outlier points
  ) +
  coord_flip() +
  labs(
    x = "Species",
    y = "Leaf mass per area (g/m<sup>2</sup>)") +
  scale_y_continuous(
    breaks = c(0, 100, 200, 300, 400),
    labels = c(0, 100, 200, 300, 400),
    limits = c(0, 400),
    expand = c(0,0)) +
  pilot::scale_color_pilot() +
  pilot::scale_fill_pilot() +
  pilot::theme_pilot(
    grid = "",
    axes = "b") +
  theme(
    legend.position = "none",
    axis.title.x = ggtext::element_markdown(),
    axis.text.y = element_text(face = "italic"))
```


Plot the species distributions of these six species with ALA occurrence data (using `galah`)
```{r, eval = F}
# install.packages(c("galah", "sf", "ozmaps"))
library(galah)
library(sf)
library(ozmaps)

# Configurate `galah` to use an email that has been registered with the ALA (https://auth.ala.org.au/userdetails/registration/createAccount)
#galah_config(email = "XXX@gmail.com", verbose = FALSE)

# Download data
#plants <- galah_call() %>%
#  galah_identify(sample_names) %>%
#  galah_apply_profile(ALA) %>%
#  atlas_occurrences()
plants <- readr::read_csv("data/plants.csv")

# Recategorise subspecies into species categories
plants <- plants %>%
  drop_na(decimalLatitude, decimalLatitude) %>%
  mutate(names = case_when(
    str_detect(scientificName, "Eucalyptus wandoo") ~ "Eucalyptus wandoo",
    str_detect(scientificName, "Pentameris airoides") ~ "Pentameris airoides",
    str_detect(scientificName, "Melaleuca armillaris") ~ "Melaleuca armillaris",
    str_detect(scientificName, "Pteridium esculentum") ~ "Pteridium esculentum",
    .default = scientificName)
  )

# Join median LMAs for each species to `plants` tibble
plants_lma <- leaf_mass_sample %>%
  group_by(taxon_name) %>%
  summarise(median_lma = median(value) %>% round(1)) %>%
  right_join(plants, by = join_by(taxon_name == scientificName)) %>%
  rename(scientificName = taxon_name) %>%
  drop_na(median_lma) # Remove NAs for unmatched subspecies

# Australia map
aus <- ozmaps::ozmap_country %>%
  st_transform(crs = st_crs(4326))

# Map points
ggplot() +
  geom_sf(
    data = aus,
    colour = "grey60",
    fill = NA) +
  geom_point(
    data = plants_lma,
    aes(x = decimalLongitude,
        y = decimalLatitude,
        colour = names),
    shape = 16,
    alpha = 0.4) +
  pilot::scale_color_pilot() +
  pilot::theme_pilot() +
  coord_sf(
    xlim = c(110, 155),
    ylim = c(-45, -10)) +
  facet_wrap(~ names, ncol = 3) +
  geom_text(
    data = plants_lma,
    mapping = aes(x = 116, y = -11,
                  label = glue::glue("LMA = {median_lma}"),
                  group = names),
    colour = "grey40",
    family = theme_get()$text$family, # use theme settings
    size = 3.5,
    lineheight = 0.92) +
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.border = element_rect(
      linewidth = 1,
      colour = "grey90",
      fill = NA)
  )
```